Cultural Transmission Can Explain the Evolution of Cooperation

### 2020-08-15 ms/main.tex
Yoav and Marc comments

If $\alpha=1$, then $\tilde{p}'= \tilde{p}$ is reduced to
\begin{equation}
\tilde{p}(1-\tilde{p})\big[(1+b)T_A + (1-c)(1-T_B)-1\big] = 0,
\end{equation}
and there are no additional equilibria.
For cooperation to take over the population (for $\tilde{p}=1$ to be globally stable) we require $\tilde{p}'>\tilde{p}$; that is,
\begin{equation}
  \tilde{p}^2 (1+b-c) + \tilde{p}(1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > \bar{w}\tilde{p} .
\end{equation}
We divide by $\tilde{p}$, set $\bar{w} = 1 + \tilde{p}(b-c)$, and rearrange to get
\begin{equation}
  (1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > 1 -\tilde{p} .
\end{equation}
Dividing by $(1-\tilde{p})$ we find that $\tilde{p}'>\tilde{p}$ if 
\begin{equation} \label{eq:vert_hori_alpha1_condition_proof}
  (1-c) (1 - T_B) + (1+b)T_A
  > 1
\end{equation}
\\

If $\alpha<1$, divide both sides of Eq.\ \ref{eq:nextgen_parents_vertical_only} by $\tilde{p}$ and set $\bar{w} = 1 + \tilde{p}(b-c)$. Then $\tilde{p}'>\tilde{p}$ if
\begin{equation}
\begin{aligned} 
  1 + \tilde{p}(b-c) < 
  & \, \tilde{p}(1+b-c) (1 - (1-\tilde{p}) (1-\alpha) T_B) \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A .
\end{aligned}
\end{equation}
Rearranging, we get
\begin{equation} 
\begin{aligned} 
  1 - \tilde{p} < 
  & - \tilde{p}(1+b-c)(1-\tilde{p}) (1-\alpha) T_B \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A .
\end{aligned}
\end{equation}
Diving by $(1-\tilde{p})$ and rearranging so that free terms are on the left and terms with $\tilde{p}$ are on the right, we have
\begin{equation} 
\begin{aligned} 
  &1 - (1-\alpha) T_A - (1+b) \alpha T_A - (1 - T_B)(1-c)  < \\
   &\tilde{p}[ - (1+b-c) (1-\alpha) T_B 
   + (1-c) (1-\alpha) T_B
   + (1+b) (1-\alpha) T_A 
   - (1-\alpha) T_A].
\end{aligned}
\end{equation}
Simplifying, we find that $\tilde{p}'>\tilde{p}$ if and only if
\begin{equation} \label{eq:vert_hori_global_condition}
c(1-T_B) - b \alpha T_A - (T_A-T_B) < \tilde{p} \cdot b (1-\alpha) (T_A-T_B).
\end{equation}

@@

We summarize these findings in the following result and corollaries.\\

\begin{result}[Vertical and horizontal transmission of cooperation] \label{result:vert_hori}
Without oblique transmission ($v=1$), fixation, extinction, and coexistence of both phenotypes are possible.
\end{result}

@@

In this case $0<v<1$, and the recursion system is more complex,
and we focus on local rather than on global stability.
To proceed, we note that 
Eq.\ \ref{eq:horizontal} can give $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$.
%Equation 4 does not give $\hat{p}$---MF
Eq.\ \ref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\tilde{p}$, since $\hat{p}$ is given in Eq.\ \ref{eq:horizontal} as a function of $\tilde{p}$ and 
Eq.\ \ref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$ (shown in Appendix~\autoref{sec:appendixB}).
We then determine the equilibria, namely, solutions of $\hat{p}' = \hat{p}$, and analyse their local stability.
%: an equilibrium $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})=\bar{w}(\hat{p}'-\hat{p})$ at the equilibrium is negative, $f'(\hat{p}^*)<0$.

@@

%We now turn to the general case where $T_A \neq T_B$. 
We apply Eqs.\ \ref{eq:horizontal}, \ref{eq:nextgen_adults_slimpify}, and \ref{eq:nextgen_parents_simplified} to obtain the function $f(\hat{p})$ (see Appendix~\autoref{sec:appendixB}):

@@

If $T=T_A=T_B$ then $\beta_1=0$ and $\beta_3=-\beta_2=\alpha b vT -cv(1-T)$, 
and $f(\hat{p})$ becomes a quadratic polynomial:
\begin{equation} \label{eq:equal_horizontal_transmission}
  f(\hat{p}) = \hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big].
\end{equation}
Clearly the only two equilibria are the fixations  $\hat{p} =  0$ and $\hat{p} = 1$.
These equilibria are locally stable if $f'(\hat{p})<0$ near the equilibrium (Appendix~\autoref{sec:appendixC}), and
\begin{equation}
f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big],
\end{equation}
with
\begin{equation} \label{eq:derivative_of_phattag-phat}
\begin{aligned}
	f'(0) &=	\alpha bvT - cv(1-T), \\
	f'(1) &=	-\alpha bvT + cv(1-T).
\end{aligned}
\end{equation}
Therefore with symmetric horizontal transmission, fixation of the cooperative phenotype ($\hat{p}=1$) occurs under the same condition as Corollary 1.1, namely Eq.\ \ref{eq:equal_transmission}.

@@

\begin{result}[Vertical, oblique, and horizontal transmission of cooperation] \label{result:vert_obli_hori}
The cultural evolution of a cooperator phenotype will follow one of the following scenarios, depending on the horizontal transmission bias $T_A-T_B$ and the coefficients $\beta_1$ and $\beta_3$:
\begin{enumerate}
\item \emph{Fixation of cooperation}: 
if \emph{(i)} $T=T_A=T_B$ and $c < b\cdot \frac{\alpha T}{1-T}$; or
\emph{(ii)} if $T_A>T_B$ and $0<\beta_3$; or 
\emph{(iii)} if $T_A<T_B$ and $\beta_1<\beta_3$.
\item \emph{Fixation of the defection}: 
if \emph{(iv)}  $T=T_A=T_B$ and $c > b\cdot \frac{\alpha T}{1-T}$; or 
\emph{(v)} if $T_A>T_B$ and $\beta_1<\beta_3<0$; or 
\emph{(vi)} if $T_A<T_B$ and $\beta_3<0$.
\item \emph{Coexistence of both phenotypes at $\hat{p}^*$}: 
if \emph{(vi)} $T_A < T_B$ and $0<\beta_3<\beta_1$.
\item \emph{Fixation of either phenotype depending on initial frequency}: if \emph{(vii)}  $T_A>T_B$ and $\beta_3<\beta_1$.
\end{enumerate}
\end{result}

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Discussion
\section*{Discussion}
We hypothesized that non-vertical transmission can explain the evolution of cooperation and evaluated this hypothesis using a deterministic discrete-time evolutionary model with fitnesses in the form of payoffs from a prisoner's dilemma game..
Under oblique and horizontal transmission, horizontal transmission bias for the cooperative phenotype was found to be sufficient and necessary for evolution of cooperation (Result~\autoref{result:obli_hori}).
Under horizontal and vertical cultural transmission, cooperation, or defection can fix, or coexist at a stable polymorphism, depending on the relationship between the cost and benefit of cooperation, the horizontal bias, and the correlation between cooperation and transmission (Result~\autoref{result:vert_hori}).
Under a combination of vertical, oblique, and horizontal transmission the dynamics are much more complicated. However, we show that under some conditions cooperation can evolve, and can even be maintained in  stable coexistence with defection (Result~\autoref{result:vert_obli_hori}).

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{} \label{sec:appendixB}
We want to find the frequency of juveniles with phenotype $A$ in next generation $\hat{p}'$ as a function of frequency
of juveniles with phenotype $A$ in the current generation $\hat{p}$.
Starting from \autoref{eq:horizontal},
\begin{equation}\label{eq:appendix_b_1}
  \hat{p}' = v \tilde{p}' + (1-v) p',
\end{equation}
we substitute $p'$ using \autoref{eq:nextgen_adults_slimpify} and $\tilde{p}'$ using \autoref{eq:nextgen_parents_simplified}, we have
\begin{equation}\label{eq:appendix_b_2}
  \begin{aligned}
  \hat{p}'  = & \frac{v}{\bar{w}}\left\{\hat{p}^2(1+b-c)\left[1-(1-\hat{p})(1-\alpha)T_B)\right]\right\} \\
  & + \frac{v}{\bar{w}}\left\{ \hat{p}(1-\hat{p})(1-c)\left[ \hat{p}(1-\alpha)T_B + 1 - T_B \right] \right\} \\
  & + \frac{v}{\bar{w}}\left\{ \hat{p}(1-\hat{p})(1+b)\left[\hat{p}(1-\alpha) + \alpha \right]T_A \right\} \\
  & + \frac{v}{\bar{w}}(1-\hat{p})^2\hat{p}(1-\alpha)T_A \\
  & + (1-v)\hat{p}^2(T_B-T_A) + (1-v)\hat{p}(1+T_A-T_B),
\end{aligned}
\end{equation}
where $\bar{w} = 1 + \hat{p}(b-c)$. 
We define $f(\hat{p})$ as
\begin{equation} \label{eq:appendix_define_f}
\begin{aligned}
      f(\hat{p}) &= \bar{w}(\hat{p}' - \hat{p})
\end{aligned}
\end{equation}
Using \emph{SymPy} \citep{Meurer2017}, a Python library for symbolic mathematics, we simplify \autoref{eq:appendix_define_f} to 
eqs.~\ref{eq:general_case_polynomial}-\ref{eq:polynomial_coefficients}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{} \label{sec:appendixC}

@@

If $p<p^*=1$, we want a condition for $1-p' < 1-p$, or
$\frac{1-p'}{1-p}<1$, or
$\lambda \frac{-(p'-p)}{1-p} < 0$, or
$-\frac{f(p)}{1-p} < 0$.
Using a linear approximation for $f(p)$ near $1$, we have
\begin{equation}\begin{aligned}
1-p' &< 1-p  \Leftrightarrow \\
\frac{f'(1)(p-1) + O\big((p-1)^2\big)}{p-1} &< 0 \Leftrightarrow \\
f'(1) - O(1-p) &< 0.
\end{aligned}\end{equation}
Therefore, if $f'(1)<0$ then there exists $\epsilon>0$ such that for any $1-\epsilon<1-p<1$ it is guaranteed that $1-p'<1-p$, that is, $p'$ is closer than $p$ to one.

\end{appendices}

>>>>

If $\alpha=1$, then $\tilde{p}'= \tilde{p}$ is reduced to
\begin{equation}
\tilde{p}(1-\tilde{p})\big[(1+b)T_A + (1-c)(1-T_B)-1\big] = 0,
\end{equation}
and there are no additional equilibria.
For cooperation to take over the population (for $\tilde{p}=1$ to be globally stable) we require $\tilde{p}'>\tilde{p}$; that is,
\begin{equation}
  \tilde{p}^2 (1+b-c) + \tilde{p}(1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
  > \bar{w}\tilde{p} .
\end{equation}
Factoring out $\tilde{p}(1-\tilde{p})$ and setting $\bar{w} = 1 + \tilde{p}(b-c)$, we find that $\hat{p}'>\hat{p}$ if
%\begin{equation}
%  (1-\tilde{p}) \big[(1-c) (1 - T_B) + (1+b)T_A\big] 
%  > 1 -\tilde{p} .
%\end{equation}
%Dividing by $(1-\tilde{p})$ we find that $\tilde{p}'>\tilde{p}$ if 
\begin{equation} \label{eq:vert_hori_alpha1_condition_proof}
  (1-c) (1 - T_B) + (1+b)T_A
  > 1
\end{equation}
\\

If $\alpha<1$, divide both sides of Eq.\ \ref{eq:nextgen_parents_vertical_only} by $\tilde{p}$ and set $\bar{w} = 1 + \tilde{p}(b-c)$. Then $\tilde{p}'>\tilde{p}$ if
\begin{equation}
\begin{aligned} 
  1 + \tilde{p}(b-c) < 
  & \, \tilde{p}(1+b-c) (1 - (1-\tilde{p}) (1-\alpha) T_B) \\
  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
  & + (1-\tilde{p})^2 (1-\alpha) T_A .
\end{aligned}
\end{equation}
%Rearranging, we get
%\begin{equation} 
%\begin{aligned} 
%  1 - \tilde{p} < 
%  & - \tilde{p}(1+b-c)(1-\tilde{p}) (1-\alpha) T_B \\
%  & + (1-\tilde{p}) (1-c) (\tilde{p} (1-\alpha) T_B + 1 - T_B) \\
%  & + (1-\tilde{p}) (1+b) (\tilde{p} (1-\alpha) + \alpha) T_A \\
%  & + (1-\tilde{p})^2 (1-\alpha) T_A .
%\end{aligned}
%\end{equation}
%Diving by $(1-\tilde{p})$ and rearranging so that free terms are on the left and terms with $\tilde{p}$ are on the right, we have
%\begin{equation} 
%\begin{aligned} 
%  &1 - (1-\alpha) T_A - (1+b) \alpha T_A - (1 - T_B)(1-c)  < \\
%   &\tilde{p}[ - (1+b-c) (1-\alpha) T_B 
%   + (1-c) (1-\alpha) T_B
%   + (1+b) (1-\alpha) T_A 
%   - (1-\alpha) T_A].
%\end{aligned}
%\end{equation}
Simplifying, we find that $\tilde{p}'>\tilde{p}$ if and only if
\begin{equation} \label{eq:vert_hori_global_condition}
c(1-T_B) - b \alpha T_A - (T_A-T_B) < \tilde{p} \cdot b (1-\alpha) (T_A-T_B).
\end{equation}

@@

We summarize these findings in the following result and corollaries.\\
%TODO:format like result 3
\begin{result}[Vertical and horizontal transmission of cooperation] \label{result:vert_hori}
Without oblique transmission ($v=1$), fixation, extinction, and coexistence of both phenotypes are possible.
\end{result}

@@

In this case $0<v<1$, and the recursion system is more complex,
and we focus on local rather than on global stability.
To proceed, we note that 
Eq.\ \ref{eq:horizontal} can give $\hat{p}'$ as a function of both $p'$ and $\tilde{p}'$.
%Equation 4 does not give $\hat{p}$---MF
Eq.\ \ref{eq:nextgen_adults_slimpify} gives $p'$ as a function of $\tilde{p}$, since $\hat{p}$ is given in Eq.\ \ref{eq:horizontal} as a function of $\tilde{p}$ and 
Eq.\ \ref{eq:nextgen_parents_simplified} gives $\tilde{p}'$ as a function of $\hat{p}$. 
Combining these equations, we find an equation for $\hat{p}'$ as a function of $\hat{p}$ (shown in Appendix~\autoref{sec:appendixA}).
We then determine the equilibria, namely, solutions of $\hat{p}' = \hat{p}$, and analyse their local stability.
%: an equilibrium $\hat{p}^*$ is locally stable when the derivative of $f(\hat{p})=\bar{w}(\hat{p}'-\hat{p})$ at the equilibrium is negative, $f'(\hat{p}^*)<0$.

@@

%We now turn to the general case where $T_A \neq T_B$. 
We apply Eqs.\ \ref{eq:horizontal}, \ref{eq:nextgen_adults_slimpify}, and \ref{eq:nextgen_parents_simplified} to obtain the function $f(\hat{p})$ (see Appendix~\autoref{sec:appendixA}):

@@

If $T=T_A=T_B$ then $\beta_1=0$ and $\beta_3=-\beta_2=\alpha b vT -cv(1-T)$, 
and $f(\hat{p})$ becomes a quadratic polynomial:
\begin{equation} \label{eq:equal_horizontal_transmission}
  f(\hat{p}) = \hat{p}(1-\hat{p})\big[\alpha bvT - cv(1-T)\big].
\end{equation}
Clearly the only two equilibria are the fixations  $\hat{p} =  0$ and $\hat{p} = 1$.
These equilibria are locally stable if $f'(\hat{p})<0$ near the equilibrium (Appendix~\autoref{sec:appendixB}), and
\begin{equation}
f'(\hat{p})=(1-2\hat{p})\big[\alpha bvT - cv(1-T)\big],
\end{equation}
with
\begin{equation} \label{eq:derivative_of_phattag-phat}
\begin{aligned}
	f'(0) &=	\alpha bvT - cv(1-T), \\
	f'(1) &=	-\alpha bvT + cv(1-T).
\end{aligned}
\end{equation}
Therefore with symmetric horizontal transmission, fixation of the cooperative phenotype ($\hat{p}=1$) occurs under the same condition as Corollary 1.1, namely Eq.\ \ref{eq:equal_transmission}.

@@

\begin{result}[Vertical, oblique, and horizontal transmission of cooperation] \label{result:vert_obli_hori}
The cultural evolution of a cooperator phenotype will follow one of the following scenarios, depending on the horizontal transmission bias $T_A-T_B$ and the coefficients $\beta_1$ and $\beta_3$:
\begin{enumerate}
\item \emph{Fixation of cooperation}: 
if \emph{(i)} $T=T_A=T_B$ and $c < b\cdot \frac{\alpha T}{1-T}$; or
if \emph{(ii)} $T_A>T_B$ and $0<\beta_3$; or 
if \emph{(iii)} $T_A<T_B$ and $\beta_1<\beta_3$.
\item \emph{Fixation of the defection}: 
if \emph{(iv)}  $T=T_A=T_B$ and $c > b\cdot \frac{\alpha T}{1-T}$; or 
if \emph{(v)} $T_A>T_B$ and $\beta_1<\beta_3<0$; or 
if \emph{(vi)} $T_A<T_B$ and $\beta_3<0$.
\item \emph{Coexistence of both phenotypes at $\hat{p}^*$}: 
if \emph{(vi)} $T_A < T_B$ and $0<\beta_3<\beta_1$.
\item \emph{Fixation of either phenotype depending on initial frequency}: if \emph{(vii)}  $T_A>T_B$ and $\beta_3<\beta_1$.
\end{enumerate}
\end{result}

%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Discussion
\section*{Discussion}
We hypothesized that non-vertical transmission can explain the evolution of cooperation and evaluated this hypothesis using a deterministic discrete-time evolutionary model with fitnesses in the form of payoffs from a prisoner's dilemma game.
Under oblique and horizontal transmission, horizontal transmission bias for the cooperative phenotype was found to be sufficient and necessary for evolution of cooperation (Result~\autoref{result:obli_hori}).
Under horizontal and vertical cultural transmission, cooperation, or defection can fix, or coexist at a stable polymorphism, depending on the relationship between the cost and benefit of cooperation, the horizontal bias, and the correlation between cooperation and transmission (Result~\autoref{result:vert_hori}).
Under a combination of vertical, oblique, and horizontal transmission the dynamics are much more complicated. However, we show that under some conditions cooperation can evolve, and can even be maintained in stable coexistence with defection (Result~\autoref{result:vert_obli_hori}).
We saw that it is likely to find configuration of parameters that results coexistence as can be seen in \autoref{fig:results2}a. In figure \autoref{fig:results2}a the yellow area in which coexistence occur has horizontal bias that favor defection ($T_B>T_A$) and the cost is relatively small.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{} \label{sec:appendixA}
We want to find the frequency of juveniles with phenotype $A$ in next generation $\hat{p}'$ as a function of frequency
of juveniles with phenotype $A$ in the current generation $\hat{p}$.
Starting from \autoref{eq:horizontal},
\begin{equation}\label{eq:appendix_b_1}
  \hat{p}' = v \tilde{p}' + (1-v) p',
\end{equation}
we substitute $p'$ using \autoref{eq:nextgen_adults_slimpify} and $\tilde{p}'$ using \autoref{eq:nextgen_parents_simplified}, we have
\begin{equation}\label{eq:appendix_b_2}
  \begin{aligned}
  \hat{p}'  = & \frac{v}{\bar{w}}\left\{\hat{p}^2(1+b-c)\left[1-(1-\hat{p})(1-\alpha)T_B)\right]\right\} \\
  & + \frac{v}{\bar{w}}\left\{ \hat{p}(1-\hat{p})(1-c)\left[ \hat{p}(1-\alpha)T_B + 1 - T_B \right] \right\} \\
  & + \frac{v}{\bar{w}}\left\{ \hat{p}(1-\hat{p})(1+b)\left[\hat{p}(1-\alpha) + \alpha \right]T_A \right\} \\
  & + \frac{v}{\bar{w}}(1-\hat{p})^2\hat{p}(1-\alpha)T_A \\
  & + (1-v)\hat{p}^2(T_B-T_A) + (1-v)\hat{p}(1+T_A-T_B),
\end{aligned}
\end{equation}
where $\bar{w} = 1 + \hat{p}(b-c)$. 
We define $f(\hat{p})$ as
\begin{equation} \label{eq:appendix_define_f}
\begin{aligned}
      f(\hat{p}) &= \bar{w}(\hat{p}' - \hat{p})
\end{aligned}
\end{equation}
Using \emph{SymPy} \citep{Meurer2017}, a Python library for symbolic mathematics, we simplify \autoref{eq:appendix_define_f} to 
eqs.~\ref{eq:general_case_polynomial}-\ref{eq:polynomial_coefficients}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{} \label{sec:appendixB}

@@

If $p<p^*=1$, we want a condition for $1-p' < 1-p$, or
$\frac{1-p'}{1-p}<1$, or
$\lambda \frac{-(p'-p)}{1-p} < 0$, or
$-\frac{f(p)}{1-p} < 0$.
Using a linear approximation for $f(p)$ near $1$, we have
\begin{equation}\begin{aligned}
1-p' &< 1-p  \Leftrightarrow \\
\frac{f'(1)(p-1) + O\big((p-1)^2\big)}{p-1} &< 0 \Leftrightarrow \\
f'(1) - O(1-p) &< 0.
\end{aligned}\end{equation}
Therefore, if $f'(1)<0$ then there exists $\epsilon>0$ such that for any $1-\epsilon<1-p<1$ it is guaranteed that $1-p'<1-p$, that is, $p'$ is closer than $p$ to one.
\end{appendices}
